NumPy---Array Programming in Python

### 2020-02-03 body.tex
Update Fig 1 and 3
Minor edits

% KJM: I moved this here to get the placement right 
% RG: minor comment on the example code: would be better to not copy IPython
%     oddities but align output on brackets and In/Out on ":"
% KJM: agreed
% FIXME we should figure out if we want to keep working on the figures in this format or not
\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array-concepts}   
  \caption{\textbf{Fundamental Array Concepts.}
    \textbf{a,} The NumPy array data structure and its associated metadata fields.
    \textbf{b,} Indexing an array with various types of arguments.
    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
    \textbf{d,} Vectorization efficiently applies operations to groups of elements
    \textbf{e,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
    \textbf{f,} Reduction operations act along one or more axes. In this
    example, a three-dimensional array is shown to be summed along various single
    axes to produce two-dimensional results, or along two axes consecutively to
    produce a one-dimensional result.
    \textbf{g,} Example NumPy code, illustrating some of these concepts.
   }
  \label{fig:array-concepts}
\end{figure*}

@@

SciPy and Matplotlib are tightly coupled with NumPy---in terms of
history, development, and use.
SciPy provides fundamental algorithms for scientific computing
including mathematical, scientific, and engineering routines.
Matplotlib generates publication-ready figures and visualizations.
The combination of NumPy, SciPy, and Matplotlib together with
an advanced interactive environment like IPython \cite{perez2007ipython}
or Jupyter \cite{Kluyver:2016aa}
provides a solid foundation for array-based scientific computing.
The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this foundation to provide several, widely used \emph{technique-specific}
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
that in turn underlay numerous \emph{domain-specific} projects
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,sunpy2015,2018EGUGA..2012146H}.
NumPy, as the focal point for the ecosystem of array-aware libraries,
sets documentation standards, and provides array testing infrastructure
and build support for integrating with Fortran code and using specialized
compilers and computing platforms.

The interactive environment created by this array programming
foundation---along with the surrounding ecosystem of tools---inside of
IPython or Jupyter, is ideally suited to exploratory data analysis.
Users fluidly inspect, manipulate, and visualize their data, and
rapidly iterate to refine programming statements. These statements are
then stitched together into imperative or functional programs, or
narrative computational notebooks.  This rich and productive
environment has made Python popular
for scientific research.

% RG: (to discuss) this figure plus associated text says that the layers about
%     Foundation build on top of that whole foundation. In reality that's not
%     true - everything builds on NumPy, almost everything on SciPy, and very
%     little on Matplotlib/IPython/Jupyter. Perhaps "interactive environment"
%     should get its own section instead.
%     Also, I think Pandas can be added next to statsmodels as statistics. That
%     avoids explaining dataframes; now it looks like a serious omission.
%     I'd also be in favor of adding Spyder in the text; scientific computing
%     beyond exploratory work is better done in an IDE.
% KJM: agreed
% FIXME we should figure out if we want to keep working on the figures in this format or not
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{static/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights just a few of the essential tools and libraries in the vast
   scientific Python ecosystem.  Most of the ecosystem builds on NumPy arrays to
   provide a rich and powerful collection of array-aware functionality.
  }
  \label{fig:ecosystem}
\end{figure}

@@

  \caption{\textbf{NumPy Array Protocol.} Through its array protocol, the NumPy API
     can operate on various types of array objects, including those written outside
     of NumPy. In this example, NumPy's \code{sum} function is called on a NumPy and
     a Dask array, respectively. The calls succeed by dispatching to the
     appropriate library implementation, and result in different output array
     types.
  }\label{fig:duck-arrays}

@@

% RG: this first sentence is duplicate with (b) three paragraphs up - on
%     purpose or not?
% KJM: yes, sort of, (a) corresponds to the two para above
% (b) corresponds to the next two para and
% (c) corresponds to the last two (or three) para
NumPy is a community of practice with a strong \emph{culture} of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project, but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

>>>>

% KJM: I moved this here to get the placement right 
\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array-concepts}   
  \caption{\textbf{Fundamental Array Concepts.}
    \textbf{a,} The NumPy array data structure and its associated metadata fields.
    \textbf{b,} Indexing an array with various types of arguments.
    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
    \textbf{d,} Vectorization efficiently applies operations to groups of elements
    \textbf{e,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
    \textbf{f,} Reduction operations act along one or more axes. In this
    example, a three-dimensional array is shown to be summed along various single
    axes to produce two-dimensional results, or along two axes consecutively to
    produce a one-dimensional result.
    \textbf{g,} Example NumPy code, illustrating some of these concepts.
   }
  \label{fig:array-concepts}
\end{figure*}

@@

SciPy and Matplotlib are tightly coupled with NumPy---in terms of
history, development, and use.
SciPy provides fundamental algorithms for scientific computing
including mathematical, scientific, and engineering routines.
Matplotlib generates publication-ready figures and visualizations.
The combination of NumPy, SciPy, and Matplotlib together with
an advanced interactive environment like IPython \cite{perez2007ipython},
or Jupyter \cite{Kluyver:2016aa} provides a solid foundation for array
programming in Python.
The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this foundation to provide several, widely used \emph{technique-specific}
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
that in turn underlay numerous \emph{domain-specific} projects
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,sunpy2015,2018EGUGA..2012146H}.
NumPy, as the focal point for the ecosystem of array-aware libraries,
sets documentation standards, and provides array testing infrastructure
and build support for integrating with Fortran code and using specialized
compilers and computing platforms.

The interactive environment created by this array programming
foundation---along with the surrounding ecosystem of tools---inside of
IPython or Jupyter, is ideally suited to exploratory data analysis.
Users fluidly inspect, manipulate, and visualize their data, and
rapidly iterate to refine programming statements. These statements are
then stitched together into imperative or functional programs, or
narrative computational notebooks.
Scientific computing beyond exploratory work is often done in a text editor
or an integrated development environment (IDEs) such as Spyder.
This rich and productive environment has made Python popular
for scientific research.

% RG: (to discuss) this figure plus associated text says that the layers about
%     Foundation build on top of that whole foundation. In reality that's not
%     true - everything builds on NumPy, almost everything on SciPy, and very
%     little on Matplotlib/IPython/Jupyter. Perhaps "interactive environment"
%     should get its own section instead.
%     Also, I think Pandas can be added next to statsmodels as statistics. That
%     avoids explaining dataframes; now it looks like a serious omission.
%     I'd also be in favor of adding Spyder in the text; scientific computing
%     beyond exploratory work is better done in an IDE.
% KJM: agreed
% FIXME we should figure out if we want to keep working on the figures in this format or not
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{static/sketches/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights just a few of the essential tools and libraries in the vast
   scientific Python ecosystem.  Most of the ecosystem builds on NumPy arrays to
   provide a rich and powerful collection of array-aware functionality.
  }
  \label{fig:ecosystem}
\end{figure}

@@

  \caption{\textbf{NumPy API and Array Protocol.} Through its array protocol, the NumPy API
     can operate on various types of array objects, including those written outside
     of NumPy. In this example, NumPy's \code{mean} function is called on a a Dask
     array.  The call succeeds by dispatching to the appropriate library implementation
     (i.e., Dask in this case) and results in a new Dask array.  Compare this
     code to the example code in Fig.~\ref{fig:array-concepts}g.
  }\label{fig:duck-arrays}

@@

NumPy is a community of practice with a strong \emph{culture} of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project, but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.
